Topics --- Example Sheet 2

### 2020-12-08 topics_statistical_theory/C1.tex
updaaaaaaaaaaates

Note that the strong law of large numbers immediately implies that $\hat{F_n}(x)$ converges almost surely to $F(x)$ as $n \to \infty$. However, the following stronger result states that this convergence holds uniformly in $x$:

@@

\begin{proof}
    See lecture notes. The main idea of the proof is to ``control'' $\hat{F_n}$ in a finite number of points $x_1, \dotsc, x_k$, and then deduce what happens between those points using the fact that distributions are increasing and right-continuous. On \href{https://en.wikipedia.org/wiki/Glivenko%E2%80%93Cantelli_theorem#Proof}{Wikipedia}, 
        a simplified proof can be found assuming that $F$ is continuous, which still encapsulates the main idea.  
\end{proof}

@@

For another consequence, we consider the problem of finding a confidence band for $F$. Given $\alpha \in (0, 1)$, set $\eps_n \ceq \sqrt{-\frac1{2n}\log(\alpha/2)}$. Then the DKW inequality tells us that
\[
\PP_F \qty( \sup_{x \in \RR} \abs{\hat{F_n}(x) - F(x)} > \eps_n) \leq \alpha,
\]
or equivalently, that 
\[
\PP_F\qty(\hat{F_n}(x) - \eps_n \leq F(x) \leq \hat{F_n}(x) + \eps_n \ \text{for all $x \in \RR$}) \geq 1 - \alpha. 
\]

We can say even more.

\begin{recap}
    For any distribution function $F$, its \emph{quantile function} is defined as
    \[
    F^{-1} \colon (0, 1] \to \RR \cup \qty{\infty} \colon p \mapsto \inf \qty{x \in \RR \mid F(x) \geq p}. 
    \]
    When necessary, we also define $F^{-1}(0) \ceq \sup\qty{x \in \RR \mid F(x) = 0}$.

    If $U \sim U(0, 1)$ and $X \sim F$, then for any $x \in \RR$ we have
    \[
    \PP(F^{-1}(U) \leq x) = \PP(U \leq F(x)) = F(x) = \PP(X \leq x). 
    \]
    This can be written simply as $F^{-1}(U) \deq X$. 
\end{recap}

Let $U_1, \dotsc, U_n \iid U(0, 1)$ with empirical distribution function $\hat{G_n}$, and let $X_1, \dotsc, X_n \iid F$. Then, we have
\begin{align*}
\hat{G_n}(F(x)) = \frac1n \sum_{i=1}^n \ind_\qty{U_i \leq F(x)} \deq \frac1n \sum_{i=1}^n \ind_\qty{X \leq x} = \hat{F_n}(x), 
\end{align*}
where $\deq$ means equality in distribution. It follows that
\[
\sup_{x \in \RR} \abs{\hat{F_n}(x) - F(x)} \deq \sup_{x \in \RR} \abs{\hat{G_n}(F(x)) - F(x)} \leq \sup_{t \in [0, 1]} \abs{\hat{G_n}(t) - t},
\]
with equality if $F$ is continuous. We conclude that if $F$ is continuous, the distribution of $\sup_{x \in \RR} \abs{\hat{F_n}(x) - F(x)}$ does not depend on $F$.

Other generalisations of \cref{thm:GC} include Uniform Laws of Large Numbers. Let $X, X_1, \dotsc, X_n$ be i.i.d.\ on a measurable space $(\Cal X, \Cal A)$, and $\Cal G$ a class of measurable functions on $\Cal X$. We say that $\Cal G$ satisfies a ULLN if 
\[
\sup_{g \in \Cal G} \abs{\frac1n \sum_{i=1}^n g(X_i) - \EE[g(X)]} \toas 0. 
\]
In \cref{thm:GC}, we showed that $\Cal  G = \qty{ \ind_\qty{\cdot \leq x} \mid x \in \RR}$ satisfies a ULLN.

@@

    Dividing by $\sigma$ yields
    \[
    \frac{\sqrt n \qty(\bar{X_n} - \mu)}{\sigma} \todist N(0, 1), 
    \]
    and multiplying both sides by $n$ and writing $V_i = \sum_{j=1}^i X_j$ we obtain
    \[
    \frac{V_i - \EE V_i}{\sqrt{\Var(V_i)}} \todist N(0, 1).
    \]
\end{recap}
Another extension starts with the observation that $\sqrt n \qty(\hat{F_n}(x) - F(x)) \todist N(0, \sigma^2)$, where
\[
\sigma^2 = \Var(\ind_\qty{X \leq x}) = \EE[\ind_{X \leq x}^2] - \EE[\ind_{X \leq x}]^2 = F(x) - F(x)^2 = F(x)(1 - F(x)). 
\]
This can be strengthened by considering $(\sqrt n \qty(\hat{F_n}(x) - F(x)) \todist N(0, \sigma^2) \mid x \in \RR)$ as a stochastic process.

\subsection{Order statistics and quantiles}
\begin{definition}
    Let $X_1, \dotsc, X_n \iid F \in \Cal F$. The \emph{order statistics} are the ordered samples $X_{(1)} \leq \dotsb \leq X_{(n)}$ (where the original order is preserved in case of a tie). 
\end{definition}

@@

\begin{proof}
    See example sheet 1, question 1.  
\end{proof}
\begin{definition}
    Let $X_1, \dotsc, X_n \iid F$. Then the \emph{sample quantile function} is defined as 
    \[
    \hat F_n^{-1}(p) = \inf \qty{x \in \RR \mid \hat F_n(x) \geq p}. 
    \]
\end{definition}

@@

For $p = \frac12$ for example, this proposition tells us that $\hat F_n^{-1}(p) = X_{(\ceil{n/2})}$, the median of the data. We now explore the distribution of $X_{(\ceil{np})}$.

@@

    Noting that $\EE V_n = \Var(V_n) = \ceil{np}$ we find
    \begin{align*}
    \sqrt{n} \qty(\frac{V_n}{n} - p) &= \frac{\sqrt{\ceil{np}}}{\sqrt n} \qty(\frac{V_n - \ceil{np}}{\sqrt{\ceil{np}}}) + \frac{\ceil{np} - np}{\sqrt n} \\
    &= \frac{\sqrt{\ceil{np}}}{\sqrt n}  \qty(\frac{V_n - \EE V_n}{\sqrt{\Var(V_n)}}) +  \frac{\ceil{np} - np}{\sqrt n}.
    \end{align*}
    Now, by the central limit theorem, the term between brackets converges to a standard $N(0, 1)$ distribution. The term $\sqrt{\ceil{np}}{\sqrt n}$ converges to $\sqrt p$ and the term $(\ceil{np} - np)/\sqrt n$ converges to 0, so by Slutsky's lemma, we find
    \[
    \sqrt{n}\qty(\frac{V_n}{n} - p) \todist \sqrt p N(0, 1) = N(0, p). 
    \]

    An analogous calculation shows that $\sqrt n \qty(\frac{W_n}{n} - (1-p)) \to N(0, 1 - p)$.

    Now we define $g \colon (0, \infty)^2 \to (0, \infty)$ by $g(x, y) \ceq x / (x + y)$, which is differentiable at $(p, 1 - p)$. Note that the distribution of $(V_n, W_n)$ is an $N(0, \smqty(p & 0 \\ 0 & q))$ distribution. By the delta method we find
    \begin{align*}
        \sqrt n \qty(U_{\ceil{np}} - p) &\deq \sqrt n\qty( g\qty(\frac{V_n}{n}, \frac{W_n}{n}) - g(p, q)) \\
        &\todist g'(p, 1-p) N\qty(0, \mqty(p & 0 \\ 0 & q)) \\
        &= N\qty(0, g'(p, 1-p) \mqty(p & 0 \\ 0 & q) g'(p, 1-p)\T) \\
        &= N(0, p(1-p)).                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        
    \end{align*}
\end{proof}

@@

\subsection{Concentration inequalities}
We turn our attention to concentration inequalities, with a focus on finite-sample results (instead of results that only hold for $n \to\infty$). 
\begin{definition}
    A random variable $X$ with mean 0 is called \emph{sub-Gaussian} with parameter $\sigma^2$ if
    \[
    M_X(t) = \EE(e^{tX}) \leq e^{t^2\sigma^2/2}
    \]
    for every $t \in \RR$. 
\end{definition}
Note that equality holds when $X \sim N(0, \sigma^2)$, since the MGF of an $N(\mu, \sigma^2)$ distribution is given by $t \mapsto \exp(\mu t + \sigma^2 t^2 /2)$.

@@

\begin{proposition}We consider some characterisations of sub-Gaussianity: 
    \begin{enumerate}[(a)]
        \item     Let $X$ be sub-Gaussian with parameter $\sigma^2$. Then 
        \begin{equation} \label{eq:sg_tail_bounds}
        \max\qty{\PP(X \geq x), \PP(X \leq -x)} \leq e^{-x^2/(2\sigma^2)} \quad\text{for every $x \geq 0$}.
        \end{equation}

@@

\begin{proof}
    \begin{enumerate}[(a)]
        \item We first consider $\PP(X \geq x)$. 
        By Markov's inequality, we have for all $t \in \RR$ that
        \[
        \PP(X \geq x) = \PP(e^{tX} \geq e^{tx}) \leq e^{-tX} \EE(e^{tX}) \leq e^{-tx + t^2\sigma^2/2}.  
        \]
        Since the LHS is independent of $t$, we can take the infimum over $t$ on the RHS and obtain
        \[
        \PP(X \geq x) \leq \inf_{t \in \RR} e^{-tx + t^2\sigma^2/2} = e^{-x^2/(2\sigma^2)}, 
        \]
        since the infimum of $t^2\sigma^2/2 - tx$ is attained at $t = x/\sigma^2$ (this method is called \emph{Chernoff bounding}).

@@

\item Note that $x \mapsto e^{-tx}$ is convex for every $t \in \RR$, so $\EE(e^{-tX}) \geq e^{-t \EE(X)} = e^{0} = 1$ by Jensen's inequality. Let $X'$ denote an independent copy of $X$: then $X - X'$ has a symmetric distribution, so all its odd moments vanish. Therefore we find
\begin{align*}
	\EE[e^{tX}] &\leq \EE[e^{-tX'}] \EE[e^{tX}] = \EE[e^{t(X - X')}] = \EE \sum_{q=0}^\infty \qty[\frac{t^{2q} (X - X')^{2q}}{(2q)!}] \\
	&= \sum_{q=0}^\infty \frac{t^{2q} \EE[(X - X')^{2q}]}{(2q)!} \leq \sum_{q=0}^\infty \frac{2^{2q-1} t^{2q} \qty(\EE[X^{2q}] + \EE[(X')^{2q}])}{(2q)!} \\
	&\leq \sum_{q=0}^\infty  \frac{2^{2q - 1}t^{2q} 2 q! C^{2q}}{(2q)!} =\sum_{q=0}^\infty  \frac{(2tC)^{2q}q!}{(2q)!} = \sum_{q=0}^\infty \frac{(2tC)^{2q}}{\prod_{j=1}^q (q+j)} \\
	&\leq \sum_{q=0}^\infty \frac{(2tC)^{2q}}{\prod_{j=1}^q (2j)} = \sum_{q=1}^\infty \frac{(2t^2C^2)^q}{q!} = e^{2t^2C^2}.
\end{align*}
This shows that $X$ is sub-Gaussian with parameter $4C^2$. 
    \end{enumerate}
\end{proof}

@@

\begin{theorem}[Hoeffding's inequality] \label{thm:hoeffding}
	Let $X_1, \dotsc, X_n$ be independent sub-Gaussian random variables, with $X_i$ having parameter $\sigma_i^2$. Then $\bar X$ is sub-Gaussian with parameter $\overline{\sigma^2}$. In particular, we have
	\[
	\max\qty{\PP(\bar X \geq x), \PP(\bar X \leq -x)} \leq e^{-nx^2/(2\overline{\sigma^2})}.
	\]
\end{theorem}

@@

Note that when $X$ takes values in $[a, b]$, its variance is at most $(b-a)^2$. However, when $\Var(X_i) \ll (b_i - a_i)^2$, Hoeffding's inequality can be loose (for example, when $X_i \sim \Rm{Bern}(p_i)$ with $p_i$ small). In such circumstances, Bennett's or Bernstein's inequality may give better results.

\begin{theorem}[Bennett's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X_i] = 0$, $\sigma_i^2 \ceq \Var(X_i) < \infty$, and $X_i \leq b$ for some $b > 0$. Define  $S \ceq \sum_{i=1}^n X_i$, $\nu \ceq \overline{\Var(X_i)}$ and $\phi \colon \RR \to \RR$ by $\phi(u) \ceq e^u - 1 - u = \sum_{k=2}^\infty \frac{u^k}{k!}$, then for every $t > 0$ we have
	\[
	\log\EE[e^{tS}] \leq \frac{n \nu }{b^2} \phi(bt). 
	\]

@@

\begin{proof}
	Define $g \colon \RR \to \RR$ by 
	\[
	g(u) \ceq \sum_{k=0}^\infty \frac{u^k}{(k+2)!} = \begin{cases}
		\frac{\phi(u)}{u^2}&\text{if $u \neq 0$}, \\
		\frac12 &\text{if $u = 0$}. 
	\end{cases} 
	\]
	Then one can check that $g$ is increasing on $\RR$, so 
	\[
	e^{tX_i} - 1 - t X_i = t^2 X_i^2 g(tX_i) \leq t^2 X_i^2 g(tb) = X_i^2 \frac{\phi(bt)}{b^2},
	\]
	and therefore
	\[
	e^{tX_i} \leq 1 + tX_i + X_i^2 \frac{\phi(bt)}{b^2} \implies \EE[e^{tX_i}] \leq 1 + \EE[X_i^2] \frac{\phi(bt)}{t^2} = 1 + \Var(X_i) \frac{\phi(bt)}{b^2}. 
	\]

	Hence for $t > 0$ we have
	\begin{align*}
		\log \EE[e^{tS}] &= \sum_{i=1}^n \log\EE[e^{tX_i}] \leq n \cdot \frac1n\sum_{i=1}^n \log\qty(1 + \Var(X_i) \frac{\phi(bt)}{b^2}) \\
		&\overset*\leq n \log(1 + \frac{\nu \phi(bt)}{b^2}) \overset{**}\leq  \frac{n\nu}{b^2}\phi(bt).
	\end{align*}
Here, $(*)$ follows from the fact that $\log$ is a concave function while $(**)$ follows from the fact that $\log(1+u) \leq u$ for all $u \geq 0$. This concludes the proof for the first part of the theorem.

@@

\begin{definition}
	For any $x \in \RR$ we define $x_+ \ceq \max(x, 0)$. 
\end{definition}

>>>>

Note that the strong law of large numbers immediately implies that $\hat{F_n}(x)$ converges almost surely to $F(x)$ as $n \to \infty$ for all fixed $x \in \RR$. However, the following stronger result states that this convergence holds uniformly in $x$:

@@

The main idea of the proof is to ``control'' $\hat{F_n}$ in a finite number of points $x_1, \dotsc, x_k$, and then deduce what happens between those points using the fact that distributions are increasing and right-continuous. On \href{https://en.wikipedia.org/wiki/Glivenko%E2%80%93Cantelli_theorem#Proof}{Wikipedia}, 
	a simplified proof can be found assuming that $F$ is continuous, which still encapsulates the main idea.  For the general proof, we need the following fact about quantile functions:
	\begin{recap}
		For any distribution function $F$, its \emph{quantile function} is defined as
		\[
		F^{-1} \colon (0, 1] \to \RR \cup \qty{\infty} \colon p \mapsto \inf \qty{x \in \RR \mid F(x) \geq p}. 
		\]
		Note that since $F$ is right-continuous and non-decreasing, the infimum is well-defined and may be replaced by a minimum, and therefore we always have $F(F^{-1}(p)) \geq p$.

		When necessary, we also define $F^{-1}(0) \ceq \sup\qty{x \in \RR \mid F(x) = 0}$.

	\end{recap}

\begin{proof}
    Let $\eps > 0$ and choose $k$ such that $\frac1k \leq \eps$. Now set $x_0 \ceq -\infty$ and $x_i \ceq F^{-1}(\frac ik)$. 
   Then we have 
    \[
    F(x_i-) - F(x_{i-1}) \leq \frac ik - \frac{i-1}{k} = \frac1k \leq \eps
    \]
    for all $i$. Define $X = \qty{x_1, \dotsc, x_k, x_1-, \dotsc, x_k-}$ (we abuse notation here) and 
    \[
    \Omega_{n, \eps} \ceq \qty{\max_{x \in X} \sup_{m \geq n} \abs{\hat F_m(x) - F(x)} \leq \eps}. 
    \]
    By a union bound and the strong law of large numbers we have
    \[
    \PP_F(\Omega_{n, \eps}\C) \leq \sum_{x \in X} \PP(\sup_{m \geq n} \abs{\hat F_m(x) - F(x)} > \eps) \to 0, 
    \]
    since $\hat F_m(x)$ and $\hat F_m(x-)$ are the sample averages of the random variables $\ind_{X \leq x}$ and $\ind_{X < x}$ and therefore converge almost surely to their means $F(x)$ and $F(x-)$.

    Now, fixing $x \in [x_{i-1}, x_i)$ we have for any $n \in \NN$, $m \geq n$
    \begin{align*}
    \hat F_m(x) - F(x) &\leq \hat F_m(x_i-) - F(x_{i-1}) \leq \hat F_m(x_i-) - F(x_i-) + F(x_i-) - F(x_{i-1}) \\
    &\leq \max_{x \in X} \sup_{m \geq n} \abs{\hat F_m(x) - F(x)} + \eps, 
        \end{align*}
    and analogously $F(x) - \hat F_n(x) \leq \max_{x \in X}\sup_{m \geq n} \abs{\hat F_m(x) - F(x)} + \eps$.

    Therefore, we have 
    \begin{align*}
    	\PP_F\qty(\sup_{m \geq n} \sup_{x \in \RR} \abs{\hat F_m(x) - F(x)} > 2\eps) \leq \PP_F \qty(\max_{x \in X} \sup_{m \geq n} \abs{\hat F_m(x) - F(x)} > \eps) = \PP(\Omega_{n, \eps}\C) \to 0. 
    \end{align*}

	Noting that $\eps$ was arbitrary, we conclude
	\begin{align*}
		\PP_F\qty(\sup_{x \in \RR} \abs{\hat F_n(x) - F(x)} \to 0) &= \PP_F\qty(\forall L \in \NN \ \exists n \in \NN \ \forall m \geq n : \sup_{x \in \RR} \abs{\hat F_n(x) - F(x)} \leq \frac1L) \\
		&= \PP_F \qty(\bigcap_{L=1}^\infty \bigcup_{n = 1}^\infty  \bigcap_{m = n}^\infty  \qty{\sup_{x \in \RR} \abs{\hat F_n(x) - F(x)} \leq \frac1L}) \\
		&= \lim_{L \to\infty} \lim_{n \to\infty} \PP_F \qty(\sup_{m \geq n} \sup_{x \in \RR} \abs{\hat F_m(x) - F(x)} \leq \frac 1L) = 1. 
	\end{align*}
\end{proof}

@@

\begin{consequence}
For another consequence, we consider the problem of finding a confidence band for $F$. Given $\alpha \in (0, 1)$, set $\eps_n \ceq \sqrt{-\frac1{2n}\log(\alpha/2)}$. Then the DKW inequality tells us that
\[
\PP_F \qty( \sup_{x \in \RR} \abs{\hat{F_n}(x) - F(x)} > \eps_n) \leq \alpha,
\]
or equivalently, that 
\[
\PP_F\qty(\hat{F_n}(x) - \eps_n \leq F(x) \leq \hat{F_n}(x) + \eps_n \ \text{for all $x \in \RR$}) \geq 1 - \alpha. 
\]
\end{consequence}

\begin{discussion}
Let $U_1, \dotsc, U_n \iid U(0, 1)$ with empirical distribution function $\hat{G}_n$, and let $X_1, \dotsc, X_n \iid F$. Then, we have
\begin{align*}
	\hat{G_n}(F(x)) = \frac1n \sum_{i=1}^n \ind_\qty{U_i \leq F(x)} \deq \frac1n \sum_{i=1}^n \ind_\qty{X_i \leq x} = \hat{F_n}(x), 
\end{align*}
where $\deq$ means equality in distribution. It follows that
\[
\sup_{x \in \RR} \abs{\hat{F_n}(x) - F(x)} \deq \sup_{x \in \RR} \abs{\hat{G_n}(F(x)) - F(x)} \leq \sup_{t \in [0, 1]} \abs{\hat{G_n}(t) - t},
\]
with equality if $F$ is continuous. We conclude that if $F$ is continuous, the distribution of $\sup_{x \in \RR} \abs{\hat{F_n}(x) - F(x)}$ does not depend on $F$, and that continuous functions give a ``worst-case'' scenario for $\sup_{x \in \RR} \abs{\hat F_n(x) - F(x)}$. 
\end{discussion}

\begin{discussion}
	Other generalisations of \cref{thm:GC} include Uniform Laws of Large Numbers. Let $X, X_1, \dotsc, X_n$ be i.i.d.\ on a measurable space $(\Cal X, \Cal A)$, and $\Cal G$ a class of measurable functions on $\Cal X$. We say that $\Cal G$ satisfies a ULLN if 
	\[
	\sup_{g \in \Cal G} \abs{\frac1n \sum_{i=1}^n g(X_i) - \EE[g(X)]} \toas 0. 
	\]
	In \cref{thm:GC}, we showed that $\Cal  G = \qty{ \ind_\qty{\cdot \leq x} \mid x \in \RR}$ satisfies a ULLN. 
\end{discussion}

@@

    Dividing by $\sigma$ yields
    \[
    \frac{\sqrt n \qty(\bar{X_n} - \mu)}{\sigma} \todist N(0, 1), 
    \]
    and multiplying both sides by $n$ and writing $V_i = \sum_{j=1}^i X_j$ we obtain
    \[
    \frac{V_i - \EE V_i}{\sqrt{\Var(V_i)}} \todist N(0, 1).
    \]
\end{recap}
\begin{discussion}
Another extension starts with the observation that $\sqrt n \qty(\hat{F_n}(x) - F(x)) \todist N(0, \sigma^2)$, where
\[
\sigma^2 = \Var(\ind_\qty{X \leq x}) = \EE[\ind_{X \leq x}^2] - \EE[\ind_{X \leq x}]^2 = F(x) - F(x)^2 = F(x)(1 - F(x)). 
\]
This can be strengthened by considering $\qty(\sqrt n (\hat{F_n}(x) - F(x)) : x \in \RR)$ as a stochastic process. 
\end{discussion}

\subsection{Order statistics and quantiles}
\begin{recap}

	If $U \sim U(0, 1)$ and $X \sim F$, then for any $x \in \RR$ we have
	\[
	\PP(F^{-1}(U) \leq x) = \PP(U \leq F(x)) = F(x) = \PP(X \leq x). 
	\]
	This can be written simply as $F^{-1}(U) \deq X$. 
\end{recap}
\begin{definition}
    Let $X_1, \dotsc, X_n \iid F \in \Cal F$. The \emph{order statistics} are the ordered samples $X_{(1)} \leq \dotsb \leq X_{(n)}$ (where the original order is preserved in case of a tie). 
\end{definition}

@@

\begin{proof}
    See example sheet 1, question 1.  
\end{proof}
\begin{definition}
    Let $X_1, \dotsc, X_n \iid F$. Then the \emph{sample quantile function} is defined as 
    \[
    \hat F_n^{-1}(p) = \inf \qty{x \in \RR \mid \hat F_n(x) \geq p}. 
    \]
\end{definition}
Note that the sample quantile function is the quantile function of the empirical distribution function.

@@

For example, this proposition tells us that $\hat F_n^{-1}(1/2) = X_{(\ceil{n/2})}$, the median of the data. We now explore the distribution of $X_{(\ceil{np})}$.

@@

    Noting that $\EE V_n = \Var(V_n) = \ceil{np}$ we find
    \begin{align*}
    \sqrt{n} \qty(\frac{V_n}{n} - p) &= \frac{\sqrt{\ceil{np}}}{\sqrt n} \qty(\frac{V_n - \ceil{np}}{\sqrt{\ceil{np}}}) + \frac{\ceil{np} - np}{\sqrt n} \\
    &= \frac{\sqrt{\ceil{np}}}{\sqrt n}  \qty(\frac{V_n - \EE V_n}{\sqrt{\Var(V_n)}}) +  \frac{\ceil{np} - np}{\sqrt n}.
    \end{align*}
    Now, by the central limit theorem, the term between brackets converges to a standard $N(0, 1)$ distribution. The term $\sqrt{\ceil{np}}/{\sqrt n}$ converges to $\sqrt p$ and the term $(\ceil{np} - np)/\sqrt n$ converges to 0, so by Slutsky's lemma, we find
    \[
    \sqrt{n}\qty(\frac{V_n}{n} - p) \todist \sqrt p N(0, 1) = N(0, p). 
    \]

    Letting $q \ceq 1 - p$, 
    an analogous calculation shows that $\sqrt n \qty(\frac{W_n}{n} - q) \to N(0, q)$.

    Now we define $g \colon (0, \infty)^2 \to (0, \infty)$ by $g(x, y) \ceq x / (x + y)$, which is differentiable at $(p, q)$ with derivative
    \[
    \nabla g(x, y) = \mqty[y / (x+y)^2 \\ -x / (x+y)^2] \implies \nabla g(p, q) = \mqty[ q \\ -p]. 
    \]Note that the distribution of $(V_n, W_n)$ is an $N(0, \smqty(p & 0 \\ 0 & q))$ distribution. By the delta method we find
    \begin{align*}
        \sqrt n \qty(U_{\ceil{np}} - p) &\deq \sqrt n\qty( g\qty(\frac{V_n}{n}, \frac{W_n}{n}) - g(p, q)) \\
        &\todist \nabla g(p, q)\T \cdot N\qty(0, \mqty(p & 0 \\ 0 & q)) \\
        &\deq  N\qty(0, \nabla g(p, q)\T  \mqty(p & 0 \\ 0 & q) \nabla g(p, q)) \\
        &\deq  N(0, pq),                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      
    \end{align*}
since $\nabla g(p, q)\T \mqty(p & 0 \\ 0 & q) \nabla g(p, q) = q^2 p + p^2 q = pq(p+q) = pq$.
\end{proof}

@@

\subsection{Concentration inequalities}
We turn our attention to concentration inequalities, with a focus on finite-sample results (instead of results that only hold for $n \to\infty$). 
\begin{definition}
    A random variable $X$ with mean 0 is called \emph{sub-Gaussian} with parameter $\sigma^2$ if
    \[
    M_X(t) = \EE(e^{tX}) \leq e^{t^2\sigma^2/2}
    \]
    for every $t \in \RR$. 
\end{definition}
\begin{remark}
Note that equality holds when $X \sim N(0, \sigma^2)$, since the MGF of an $N(\mu, \sigma^2)$ distribution is given by $t \mapsto \exp(\mu t + \sigma^2 t^2 /2)$. 
	content...
\end{remark}

@@

\begin{discussion}[Chernoff bounding]
	Let $X$ be any random variable, then by Markov's inequality, then we have for all $t > 0$ that 
	\[
	\PP(X \geq x) = \PP(e^{tX} \geq e^{tx}) \leq e^{-tx} \EE[e^{tX}] = e^{-tx} M_X(t).
	\]
	Since the left-hand side is independent of $t$, we can minimise over all $t$ and obtain \[
	\PP(X \geq x) \leq \inf_{t > 0} e^{-tx} M_X(t), 
	\]
	which often gives better results than the ``standard'' Markov bound. This technique is called \emph{Chernoff bounding}, and is very useful if bounds on $M_X(t)$ are known. 
\end{discussion}
\begin{proposition}We consider some characterisations of sub-Gaussianity: 
    \begin{enumerate}[(a)]
        \item     Let $X$ be sub-Gaussian with parameter $\sigma^2$. Then 
        \begin{equation} \label{eq:sg_tail_bounds}
        \max\qty{\PP(X \geq x), \PP(X \leq -x)} \leq e^{-x^2/(2\sigma^2)} \quad\text{for every $x \geq 0$}.
        \end{equation}

@@

\begin{proof}
    \begin{enumerate}[(a)]
        \item We first consider $\PP(X \geq x)$. 
        By a Chernoff bound, we have
        \[
        \PP(X \geq x) \leq \inf_{t \in \RR} e^{-tx + t^2\sigma^2/2} = e^{-x^2/(2\sigma^2)}, 
        \]
        since the infimum of $t^2\sigma^2/2 - tx$ is attained at $t = x/\sigma^2$.

@@

\item Note that $x \mapsto e^{-tx}$ is convex for every $t \in \RR$, so $\EE(e^{-tX}) \geq e^{-t \EE(X)} = e^{0} = 1$ by Jensen's inequality. Let $X'$ denote an independent copy of $X$: then $X - X'$ has a symmetric distribution, so all its odd moments vanish. Therefore we find
\begin{align*}
	\EE[e^{tX}] &\leq \EE[e^{-tX'}] \EE[e^{tX}] = \EE[e^{t(X - X')}]= \EE \sum_{q=0}^\infty \qty[\frac{t^q (X - X')^{q}}{q!}] 
	\\
	&\overset\star= \sum_{q=0}^\infty \frac{t^{2q} \EE[(X - X')^{2q}]}{(2q)!} \leq \sum_{q=0}^\infty \frac{2^{2q-1} t^{2q} \qty(\EE[X^{2q}] + \EE[(X')^{2q}])}{(2q)!} \\
	&\leq \sum_{q=0}^\infty  \frac{2^{2q - 1}t^{2q} 2 q! C^{2q}}{(2q)!} =\sum_{q=0}^\infty  \frac{(2tC)^{2q}q!}{(2q)!} = \sum_{q=0}^\infty \frac{(2tC)^{2q}}{\prod_{j=1}^q (q+j)} \\
	&\leq \sum_{q=0}^\infty \frac{(2tC)^{2q}}{\prod_{j=1}^q (2j)} = \sum_{q=1}^\infty \frac{(2t^2C^2)^q}{q!} = e^{2t^2C^2}.
\end{align*}
Here, $\star$ follows from Fubini's theorem and the fact that the odd moments of $X - X'$ vanish. 
This shows that $X$ is sub-Gaussian with parameter $4C^2$. 
    \end{enumerate}
\end{proof}

@@

\begin{theorem}[Hoeffding's inequality] \label{thm:hoeffding}
	Let $X_1, \dotsc, X_n$ be independent sub-Gaussian random variables, with $X_i$ having parameter $\sigma_i^2$. Then $\bar X$ is sub-Gaussian with parameter $\overline{\sigma^2} = \frac1n \sum_{i=1}^n \sigma_i^2$. In particular, we have
	\[
	\max\qty{\PP(\bar X \geq x), \PP(\bar X \leq -x)} \leq e^{-nx^2/(2\overline{\sigma^2})}.
	\]
\end{theorem}

@@

\begin{example}
	Note that when $X$ takes values in $[a, b]$, its variance is at most $(b-a)^2$. However, when $\Var(X_i) \ll (b_i - a_i)^2$, Hoeffding's inequality can be loose (for example, when $X_i \sim \Rm{Bern}(p_i)$ with $p_i$ small). In such circumstances, Bennett's or Bernstein's inequality may give better results (example sheet 1, question 4). 
\end{example}

\begin{theorem}[Bennett's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X_i] = 0$, $\sigma_i^2 \ceq \Var(X_i) < \infty$, and $X_i \leq b$ for some $b > 0$. Define  $S \ceq \sum_{i=1}^n X_i$, $\nu \ceq \overline{\sigma^2}$ and $\phi \colon \RR \to \RR$ by $\phi(u) \ceq e^u - 1 - u = \sum_{k=2}^\infty \frac{u^k}{k!}$, then for every $t > 0$ we have
	\[
	\log\EE[e^{tS}] \leq \frac{n \nu }{b^2} \phi(bt). 
	\]

@@

\begin{proof}
	Define $g \colon \RR \to \RR$ by 
	\[
	g(u) \ceq \sum_{k=0}^\infty \frac{u^k}{(k+2)!} = \begin{cases}
		\frac{\phi(u)}{u^2}&\text{if $u \neq 0$}, \\
		\frac12 &\text{if $u = 0$}. 
	\end{cases} 
	\]
	Then one can check that $g$ is increasing on $\RR$, so 
	\[
	e^{tX_i} - 1 - t X_i = t^2 X_i^2 g(tX_i) \leq t^2 X_i^2 g(tb) = X_i^2 \frac{\phi(bt)}{b^2},
	\]
	and therefore
	\[
	e^{tX_i} \leq 1 + tX_i + X_i^2 \frac{\phi(bt)}{b^2} \implies \EE[e^{tX_i}] \leq 1 + \EE[X_i^2] \frac{\phi(bt)}{t^2} = 1 + \sigma_i^2 \frac{\phi(bt)}{b^2}. 
	\]

	Hence for $t > 0$ we have
	\begin{align*}
		\log \EE[e^{tS}] &= \sum_{i=1}^n \log\EE[e^{tX_i}] \leq n \cdot \frac1n\sum_{i=1}^n \log\qty(1 + \sigma_i^2 \frac{\phi(bt)}{b^2}) \\
		&\overset*\leq n \log(1 + \frac{\nu \phi(bt)}{b^2}) \overset{**}\leq  \frac{n\nu}{b^2}\phi(bt).
	\end{align*}
Here, $(*)$ follows from the fact that $\log$ is a concave function while $(**)$ follows from the fact that $\log(1+u) \leq u$ for all $u \geq 0$. This concludes the proof for the first part of the theorem.

@@

%\begin{definition}
%	For any $x \in \RR$ we define $x_+ \ceq \max(x, 0)$. 
%\end{definition}
